Classical Planning Algorithms ¡ on the Atari Video Games: Preliminary Results

### 2014-11-18 icaps15-paper/paper.tex
Table and experiments changed

\newcommand{\Asterix}{\propername{Asterix}}
\newcommand{\BeamRider}{\propername{Beam Rider}}
\newcommand{\Breakout}{\propername{Breakout}}
\newcommand{\Enduro}{\propername{Enduro}}
\newcommand{\Freeway}{\propername{Freeway}}
\newcommand{\Pong}{\propername{Pong}}
\newcommand{\Qbert}{\propername{Q*Bert}}
\newcommand{\Seaquest}{\propername{Seaquest}}
\newcommand{\SpaceInvaders}{\propername{Space Invaders}}
\newcommand{\Deepmind}{\propername{Deepmind}}

@@

\end{table*}

We tested  $\IW(1)$ and \BFS\  over $9$ games: those considered in the \emph{training} set
of \cite{bellemare:jair2013}, abbreviated  as BNVB, and those discussed in \cite{deep-mind-atari}.
In Table~\ref{performance},  the performance of  $\IW(1)$ and \BFS\
% , characterized by the average score obtained over all episodes, the standard deviation
% of the score obtained and time (in seconds) per lookahead,
is compared with that of \UCT, breadth-first search (\BRFS), and random action selection.
The discount factor used by all the algorithms is $\gamma=0.995$ and the average scores
reported for \UCT\ in  BNVB  are included in  parenthesis for reference. 
Our experimental setup  follows the one in BNVB with one exception: a maximum budget of  $150,000$ simulated frames 
is applied to all the algorithms, including  \UCT\ that is limited to  $500$ base policy rollouts of depth $300$.
% \footnote{In \UCT,  $500$ base policy rollouts
% of depth $300$ is equivalent to a budget of $150,000$ simulated frames.}.
The bound on the number of simulated frames is like a bound on lookahead time, 
as most of the time in the lookahead search is spent in the calls to the emulator
for computing the next RAM state.  Also in order to save time, as in BNVB,   all of the algorithms 
reuse the subtree from  the previous lookahead that is rooted in the selected
child, deleting its siblings and their descendants. More precisely, no calls to the
emulator are done  for transitions that are cached in that subtree. As a result
the number of nodes in the lookahead tree may be larger actually than the budget.
In addition, in  \IW, the states that are ``reused''  from the  previous searches
are ignored in the computation of the novelty of new states so that more states 
can  escape pruning. Otherwise, \IW\ often doesn't use the full budget. This is not 
needed in \BFS\ which does no pruning.  $\IW(1)$ and \BFS\ have been  limited 
search up to a depth of $m=1500$  frames and up to $m=20,000$ frames per root branch.
This is to avoid the search for going too deep or being too committed to a single
root action. (**** Please revise ****).

Table~\ref{performance} shows that $\IW(1)$ and \BFS\ both outperform \BRFS\  and the random baseline.
\BRFS\ rarely collects any reward, as the depth of the \BRFS\ search tree
results in  a lookahead of $0.3$ seconds ($20$  frames or $4$ nodes deep).
On the other hand, both $\IW(1)$ and \BFS\ reach states that are  up to $350$--$1500$ frames deep
($70$--$260$ nodes or $6$--$22$ seconds), even if $\IW(1)$ doesn't always use all the simulation frames
allocated due to its agressive pruning.
%% case of \BFS, due to having a novelty of $0$ and the planning horizon being limited to $m$
%%% (*** What is novelty $0$??? Not clear why BFS doesn't take all the budget.

Interestingly, simple $\IW(1)$ outperforms \UCT\ in $4$ out of $9$ games,
both in terms of  average scores and average times per call. \BFS\ is
faster than $\IW(1)$ and an order-of-magnitude faster than \UCT\ in general,
and it  outscores \UCT\ in $2$ of the games (\Freeway\ and
\Seaquest). Compared with the \UCT\ results in   \cite{bellemare:jair2013}, shown in parenthesis,
both $\IW(1)$ and \BFS\  outperform \UCT\ in $3$ of the  $9$ games (\Breakout, \Enduro\ and \Freeway),
with \BFS\ and \UCT\ tied in  \Pong. It'd  be very interesting to see the results of
\BFS\ when given more time per action, but we don't have those results yet.
As mentioned in the previous section, the order in which $\IW(1)$ and \BFS\ generate the children
of an expanded node are randomized, something which has a potential effect on the  novelty measures
of nodes, and hence in the search graph. We found that this randomization helps, in particular
in \Qbert, where not randomizing results  in $\IW(1)$ being trapped into  loops sometimes.

We also  note that in \Freeway,  both $\IW(1)$ and \BFS\ attain a score similar to the
baseline semi-random algorithm \emph{Perturb} in  \cite{bellemare:jair2013},
which outperforms   both \UCT\ and  the reinforcement learning methods.
Likewise,  $\IW(1)$ outperforms the best learning algorithm in \cite{deep-mind-atari}
in $4$ out of the  $7$ games considered, and \BFS\ does so in $5$  of
the $7$ games.

>>>>

\newcommand{\Asterix}{\propername{Asterix}}
\newcommand{\BeamRider}{\propername{Beam Rider}}
\newcommand{\Breakout}{\propername{Breakout}}
\newcommand{\Enduro}{\propername{Enduro}}
\newcommand{\Freeway}{\propername{Freeway}}
\newcommand{\Pong}{\propername{Pong}}
\newcommand{\Qbert}{\propername{Q*Bert}}
\newcommand{\Seaquest}{\propername{Seaquest}}
\newcommand{\SpaceInvaders}{\propername{Space Invaders}}
\newcommand{\Deepmind}{\propername{Deepmind}}
\newcommand{\CrazyClimber}{\propername{Crazy\_climber}}
\newcommand{\Kangaroo}{\propername{Kangaroo}}
\newcommand{\Pooyan}{\propername{Pooyan}}
\newcommand{\Berzerk}{\propername{Berzerk}}
\newcommand{\Bowling}{\propername{Bowling}}
\newcommand{\Montezuma}{\propername{Montezuma\_revenge}}
\newcommand{\StarGunner}{\propername{Star\_gunner}}
\newcommand{\Venture}{\propername{Venture}}

@@

\end{table*}
 We tested $\IW(1)$ and \BFS\ over $35$ games: those considered in
\cite{bellemare:jair2013}, abbreviated as BNVB \footnote{The full set
  of games is made of 55 games, 5 runs each. As each run takes around
  24 hours of CPU time, we luck the resources to submit the full table
  by the submission date, but the full table will be ready for the
  camera ready version.}. In Table~\ref{performance}, the
performance of $\IW(1)$ and \BFS\
% , characterized by the average score obtained over all episodes, the standard deviation
% of the score obtained and time (in seconds) per lookahead,
is compared with that of \UCT\ and breadth-first search (\BRFS).
The discount factor used by all the algorithms is $\gamma=0.995$ and the average scores
reported for \BRFS\ and \UCT\ in  BNVB  are included for reference. 
Our experimental setup  follows the one in BNVB with one exception: a maximum budget of  $150,000$ simulated frames 
is applied to \textit{all the algorithms}, including  \UCT\ that is limited to  $500$ base policy rollouts of depth $300$.
% \footnote{In \UCT,  $500$ base policy rollouts
% of depth $300$ is equivalent to a budget of $150,000$ simulated frames.}.
The bound on the number of simulated frames is like a bound on
lookahead time, as most of the time in the lookahead search is spent
in the calls to the emulator for computing the next RAM state.  Also,
as in BNVB, all of the algorithms reuse the subtree from the previous
lookahead that is rooted in the selected child, deleting its siblings
and their descendants. More precisely, no calls to the emulator are
done for transitions that are cached in that subtree. As a result the
number of nodes in the lookahead tree may be larger actually than the
budget. In addition, in \IW, the states that are ``reused'' from the
previous searches are ignored in the computation of the novelty of new
states so that more states can escape pruning. Otherwise, \IW\ often
doesn't use the full budget. This is not needed in \BFS\ which does no
pruning.  $\IW(1)$ and \BFS\ have been limited the search to a depth of
$m=1500$ frames and up to $m=150,000$ frames per root branch. This is
to avoid the search for going too deep or being too committed to a
single root action. Similarly UCT defines a limit of $m=300$ frames
for each rollout. Note that UCT is more likely to have a performance
loss if the depth is increased, as the number of rollouts would
decrease in order to maintain the same budget of $150,000$ simulated
frames. The last difference inherited from BNVB between all algorithms
and UCT, is that UCT selects an action randomly in the rollout every
single frame, while the others select an action every 5 frames, and
the remain ones are $no-op$, as is done in the simulation
environment. Therefore, a node in all algorithms represent $5$ frames,
but in UCT represents $1$ frame. As a result, a UCT rollout of depth
$300$ frames involve the selection of 300 actions, decreasing the
budget of simulated frames by $300$, while the other algorithms need a
branch of $1500$ frames to select $300$ actions, decreasing the budget
by $1500$ frames instead.

Table~\ref{performance} shows that $\IW(1)$ and \BFS\ both outperform
\BRFS, which rarely collects any reward in many domains, as the
depth of the \BRFS\ search tree results in a lookahead of $0.3$
seconds ($20$ frames or $4$ nodes deep). On the other hand, both
$\IW(1)$ and \BFS\ reach states that are up to $350$--$1500$ frames
deep ($70$--$260$ nodes or $6$--$22$ seconds), even if $\IW(1)$
doesn't always use all the simulation frames allocated due to its
agressive pruning. This behaviour can be observed in games such as
\Breakout, \CrazyClimber, \Kangaroo, and \Pooyan, where the average
CPU time for each lookahead is up to $10$ times faster than \BFS. Note
that computation time for \UCT\ and \BRFS\ is similar to \BFS, as the
most expensive computation is to generate a frame through the
simulator and all algorithms use always the full budget.

Interestingly, in terms of average scores out of $35$ games, simple $\IW(1)$ outperforms
\BFS\ and \UCT\ in $18$ and $23$ games respectively,
most often with lower average times per call. \BFS\ outscores \UCT\ in
$22$ of the games \footnote{Note that both \IW\ and \BFS\ may improve, as 6 and 4 games  respectively are not reported in the table
  yet.}. As mentioned in the
previous section, the order in which $\IW(1)$ and \BFS\ generate the
children of an expanded node is randomized, something which has a
potential effect on the novelty measures of nodes, and hence in the
search graph. We found that this randomization helps, in particular in
\Qbert, where not randomizing results in $\IW(1)$ being trapped into
loops sometimes.

Likewise, $\IW(1)$ outperforms the best learning algorithm in
\cite{deep-mind-atari} in $7$ out of the $7$ games considered, and
\BFS\ does so in $6$ of the $7$ games.

We also note that in \Freeway, \Berzerk, \Bowling, \Montezuma,
\StarGunner, and \Venture, both $\IW(1)$ and \BFS\ attain a better
score than the baseline semi-random algorithm \emph{Perturb} in
\cite{bellemare:jair2013}, which outperforms both \UCT\ and the
reinforcement learning methods. \emph{Perturb} performance helps
understanding these domains. \emph{Perturb} selects a fixed action with probability
0.95 and otherwise acts uniformly randomly, and the best run out of 10
episodes is reported in BNVB. Note that all the other algorithms
report average score instead. The behaviour of \emph{Perturb} hinder
the nature of these domains: rewards are very deep, but tipically
repeating the same action may lead to some reward.

Consider \Freeway, where a chicken has to run across a ten lane
highway filled with traffic. Every time it gets across, we
receive a point, i.e. a reward of 1. If the chicken is hit by a car,
the chicken is forced back. 12 out of 18 possible actions have an
effect: move up ( up-right, up-left, up-fire, up-right-fire,
up-left-fire, all have same effect as up ) or move down ( down-right,
down-left, down-fire, down-right-fire, down-left-fire, all have same
effect as down ), the remaining 6 actions are like no-ops. It's easy
to understand that \emph{Perturb} would achieve a high score if the
selected action is move up with 0.95 probability. In the following link
\url{http://www.youtube.com/playlist?list=PLXpQcXUQ_CwenUazUivhXyYvjuS6KQOI0}
we can observe videos of \IW\ and \UCT\ playing \Freeway. As reported
in Table~\ref{performance}, \UCT\ does not manage to take any chicken
across the highway, while \IW\ is the best performer out of all tested
algorithms. The main reason why \UCT\ performs so erratically is that
the rollouts do not manage to collect any reward, as there is a
$3^{-1}$ probability to choose any move up action, and given that the
depth of the closest reward is about 200 frames deep, the probability
of a rollout achieveing any reward is $3^{-200}$, which implies
choosing 200 move-up actions. On the other hand, \IW\ manages to find
rewards without exhausting the full budget. The main reason that
explains this behaviour is that at most only 3 generated states from
the same parent node will pass the novelty test, as in any expanded
node there is at most 3 different new possible bytes in the RAM state
resulting from moving up, down, or do-nothing.

Another challenging game is \Montezuma. In this game rewards are very
sparse, deep, and most of the actions lead to loosing a life, which
only becomes effective after hundreds of frames, i.e. up-left may
enforce jumping and falling until we loose a life. All algorithms achieve 0 score, wheter RL-based or Planning-based. But, as
shown in Table~\ref{performance}, \BFS\ achieves an average score of
80, which is still a very low score, as the first possible reward
increases the score in 100. The best run by \BFS\ achieves a score of
400, pointing out that \BFS\ still is not able to consistently find rewards. One way to improve the performance of any algorithm in \Montezuma\ is to encode the lost of a life as a negative reward. Capturing the lost of lifes in the reward function is relevant to many games where not loosing a life is important in later stages of the game. Currently,
rewards are encoded directly as the change of score, which does not distinguish between the benefits of keeping lifes. This
beahaviour is also observed in other games like \Breakout\ and \SpaceInvaders, where \BFS\
is the best performer by far, although initially it looses all lifes but 1 and then plays extremely well. Videos on the link above illustrate
this behaviour.
